---
title: "Diffusion profiles"
date: last-modified
date-format: YYYY-MM-DD
vignette: >
  %\VignetteIndexEntry{Diffusion profiles}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author:
  Gregor Lueg
---

# Diffusion profiles

## Setup

```{r}
#| label: setup
library(genewalkR)
library(data.table)
library(magrittr)
```

## Background

[Ruiz, Zitnik and Leskovec](https://www.nature.com/articles/s41467-021-21770-8)
built the multiscale interactome: drugs and diseases hang off proteins, proteins
talk to each other and to a hierarchy of biological functions. To explain how a
drug treats a disease, they run one personalised PageRank per drug and per
disease and compare the resulting visitation vectors, the diffusion profiles.
Two drugs that hit the same biology end up with similar profiles, and a drug
whose profile looks like a disease's is a candidate treatment.

It's a personalised PageRank with two twists:

- **Sinks.** The seed is a source. Every other node of a sink type absorbs the
  walker: mass flows in, nothing flows out. In the paper drugs and diseases are
  sinks, so a drug's profile doesn't leak through some other drug into its
  targets.
- **Type weights.** At each step the walker first picks one of the node types
  among its neighbours, proportional to the type weight, then a neighbour of
  that type. This lets you push the walk towards, say, biological functions,
  no matter how many protein neighbours a node has.

Leave `type_weights = NULL` and `sink_types` empty and you get the plain
(edge-weighted) personalised PageRank. The graph is built once in Rust and the
seeds run in parallel.

## A toy interactome

Two drugs, two diseases, six proteins in a chain and three functions. `drug_A`
and `disease_X` share `P2`; `drug_B` and `disease_Y` sit at the other end of
the chain.

```{r}
#| label: toy graph
toy_nodes <- data.table(
  id = c(
    "drug_A",
    "drug_B",
    paste0("P", 1:6),
    "F1",
    "F2",
    "F3",
    "disease_X",
    "disease_Y"
  ),
  type = c(
    rep("drug", 2),
    rep("protein", 6),
    rep("function", 3),
    rep("disease", 2)
  )
)

toy_edges <- data.table(
  from = c(
    # drugs and diseases to proteins
    "drug_A", "drug_A", "drug_B", "disease_X", "disease_X", "disease_Y",
    # protein chain
    "P1", "P2", "P3", "P4", "P5",
    # proteins to functions
    "P1", "P2", "P3", "P4", "P5", "P6"
  ),
  to = c(
    "P1", "P2", "P5", "P2", "P3", "P6",
    "P2", "P3", "P4", "P5", "P6",
    "F1", "F1", "F2", "F2", "F3", "F3"
  )
)

toy_seeds <- c("drug_A", "drug_B", "disease_X", "disease_Y")
```

First the plain personalised PageRank: no sinks, no type weights.

```{r}
#| label: toy plain
plain <- DiffusionProfiles(
  graph_dt = toy_edges,
  node_dt = toy_nodes,
  sink_types = character()
) %>%
  generate_profiles(seeds = toy_seeds, .verbose = FALSE)

round(get_profiles(plain), 3)
```

Each column sums to 1. Look at `drug_A`: 0.075 of its mass sits on
`disease_X`, and from there it keeps walking into `P3` and beyond. Now make
drugs and diseases sinks and up-weight functions five-fold.

```{r}
#| label: toy constrained
constrained <- DiffusionProfiles(
  graph_dt = toy_edges,
  node_dt = toy_nodes,
  sink_types = c("drug", "disease"),
  type_weights = c(drug = 1, protein = 1, `function` = 5, disease = 1)
) %>%
  generate_profiles(seeds = toy_seeds, .verbose = FALSE)

constrained

round(get_profiles(constrained), 3)
```

`F1` goes from 0.123 to 0.289 in the `drug_A` profile, and `F3` more than
doubles in the `drug_B` one. The profiles are now dominated by the functions
each drug touches, which is exactly what you want to compare on.

Distances between profiles are computed in Rust as well. Lower means more
similar. Ruiz et al. use the correlation distance to rank drug-disease pairs,
which is the default here; `"canberra"`, `"l1"`, `"l2"` and `"cosine"` are
also available.

```{r}
#| label: toy distances
calculate_profile_distances(
  constrained,
  from = c("drug_A", "drug_B"),
  to = c("disease_X", "disease_Y")
) %>%
  round(3)
```

`drug_A` pairs with `disease_X`, `drug_B` with `disease_Y`. Good.

## Reactome and a MYC signature

The DuckDB has no drugs or diseases, but Reactome gives us genes, pathways and
the pathway hierarchy. We play the drug-disease game with gene signatures
instead: a node for the MYC target genes that ship with the package, and a node
for a random gene set of the same size as the control. Both get type
`signature`, and signatures are sinks.

```{r}
#| label: reactome graph
gene_gene <- get_interactions_reactome()[, .(
  from = as.character(from),
  to = as.character(to)
)]
gene_pathway <- get_gene_to_reactome()[, .(
  from = as.character(from),
  to = as.character(to)
)]
pathway_pathway <- get_reactome_hierarchy("child_of")[, .(
  from = as.character(from),
  to = as.character(to)
)]

all_genes <- unique(c(gene_gene$from, gene_gene$to, gene_pathway$from))

set.seed(42L)
myc_targets <- intersect(myc_genes$ensembl_gene, all_genes)
random_genes <- sample(all_genes, length(myc_targets))

signature_edges <- rbind(
  data.table(from = "MYC_targets", to = myc_targets),
  data.table(from = "random", to = random_genes)
)

edges <- rbind(gene_gene, gene_pathway, pathway_pathway, signature_edges) %>%
  unique() %>%
  .[from != to]

nodes <- rbind(
  data.table(id = all_genes, type = "gene"),
  data.table(
    id = unique(c(gene_pathway$to, pathway_pathway$from, pathway_pathway$to)),
    type = "pathway"
  ),
  data.table(id = c("MYC_targets", "random"), type = "signature")
) %>%
  unique(by = "id")

nodes[, .N, by = type]
nrow(edges)
```

For readable output, the pathway names. A few pathway ids in the hierarchy
have no name in the database; those keep their id.

```{r}
#| label: reactome labels
pathway_labels <- get_reactome_info()[, .(
  id = as.character(reactome_id),
  label = as.character(reactome_name)
)]

label_pathways <- function(ids) {
  fcoalesce(pathway_labels[match(ids, id), label], ids)
}
```

We generate a profile for both signatures and for every single pathway. That's
close to 3,000 PageRanks over more than 14k nodes.

```{r}
#| label: reactome profiles
pathways <- nodes[type == "pathway", id]

reactome_profiles <- DiffusionProfiles(
  graph_dt = edges,
  node_dt = nodes,
  sink_types = "signature"
)

tictoc <- system.time(
  reactome_profiles <- generate_profiles(
    reactome_profiles,
    seeds = c("MYC_targets", "random", pathways)
  )
)

tictoc
```

A few seconds. Blazingly fast.

### Where does the mass go?

The obvious first look: which pathways soak up the most mass from each
signature?

```{r}
#| label: top mass
profiles <- get_profiles(reactome_profiles)

top_mass <- function(seed, n = 8L) {
  mass <- sort(profiles[pathways, seed], decreasing = TRUE)[seq_len(n)]
  data.table(pathway = label_pathways(names(mass)), mass = signif(mass, 3))
}

top_mass("MYC_targets")
top_mass("random")
```

MYC targets land on mRNA splicing, rRNA processing and nucleotide metabolism.
MYC drives ribosome biogenesis and nucleotide synthesis, so that checks out.
The random set lands on whatever is big and well connected: olfactory
receptors, the generic transcription pathway, neutrophil degranulation. The
latter also shows up for MYC. Raw mass is biased towards hubs.

### Comparing profiles

This is where the approach earns its keep. Instead of asking where the
signature's walker ends up, ask which pathway's own profile looks most like
the signature's.

```{r}
#| label: distances
dists <- calculate_profile_distances(
  reactome_profiles,
  from = c("MYC_targets", "random"),
  to = pathways
)

closest <- function(seed, n = 8L) {
  d <- sort(dists[seed, ])[seq_len(n)]
  data.table(pathway = label_pathways(names(d)), distance = round(d, 3))
}

closest("MYC_targets")
closest("random")
```

The hubs are gone. For MYC you get pyrimidine and PRPP biosynthesis,
nucleotide interconversion and polyamine metabolism (ODC1 is one of the
classic MYC targets), i.e. the anabolic programme MYC is known for. The random
set gives a grab bag of small pathways without a common theme.

Don't overread the absolute numbers though. On a graph this size all the
correlation distances sit between 0.8 and 0.9, and the random set's best hit
(0.842) is barely further away than MYC's (0.826). The ranking within a
signature carries the biology; whether a single distance is surprising needs a
null, see below.

## Caveats and next steps

- Profiles are dense `nodes x seeds` matrices. 2,847 seeds over 14,295 nodes is
  about 325 MB of doubles, so thousands of seeds on a large graph get painful
  fast.
- There's no permutation test yet. For a proper null you'd rewire the graph
  (the configuration model used for GeneWalk would do) and regenerate the
  profiles. Watch this space.
- Type weights are a free parameter. Ruiz et al. tuned theirs on known
  drug-disease pairs; without such a gold standard, leave them at `NULL` or
  keep them flat.
