---
title: "Graph Diffusion"
vignette: >
  %\VignetteIndexEntry{Graph Diffusion}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author:
  Gregor Lueg
---

# Network diffusion with genewalkR

## Setup
```{r}
#| label: setup
library(genewalkR)
library(data.table)
library(igraph)
library(ggplot2)
```

## Background

Network diffusion propagates node-level signals (e.g. gene-level scores)
across a graph using a kernel derived from the graph Laplacian. The intuition
is straightforward: if a node carries a strong signal, its neighbours should
inherit some of that signal, with the strength decaying as graph distance
increases. This is useful for tasks based on the "guilty by association" 
principle, where you have a handful of known disease genes and want to find 
additional candidates that are well-connected to them in a protein-protein 
interaction network.

`genewalkR` implements five spectral kernels from the diffusion literature:

- **Regularised Laplacian**: $(\\sigma^2 L + \\alpha I)^{-1}$
- **Commute time**: Moore-Penrose pseudoinverse of $L$
- **Diffusion (heat)**: $\\exp(-\\sigma^2 / 2 \\cdot L)$
- **Inverse cosine**: $\\cos(\\pi/4 \\cdot \\lambda)$
- **p-step**: $(aI - L)^p$

And three scoring methods:

- **Raw**: direct kernel-weighted scores
- **Z-score**: analytically normalised scores under a permutation null
- **Monte Carlo (heatrank)**: empirical p-values via permutation testing

The entire computation is performed in Rust via a spectral decomposition of the
Laplacian, so the dense kernel matrix is never materialised. The work is
based on the [diffuStats package](https://pmc.ncbi.nlm.nih.gov/articles/PMC5860365/).

## A simple example: lattice graph

We start with a square lattice to build intuition. On a lattice, graph distance
is easy to reason about, so we can verify that diffusion scores decay as
expected.

### Building the graph and kernel

```{r}
#| label: lattice-setup

# 15 x 15 lattice = 225 nodes
g <- make_lattice(dimvector = c(15, 15))
V(g)$name <- as.character(seq_len(vcount(g)))

obj <- DiffusionScores(g)
obj
```

Run the kernel:

```{r}
#| label: lattice-kernel

obj <- build_kernel(
  obj,
  kernel = "regularised_laplacian",
  kernel_params = params_kernel(sigma2 = 1.0, add_diag = 1.0),
  normalised = TRUE,
  strategy = "full",
  .verbose = TRUE
)
obj
```

### Diffusing a single source node

We place a unit signal on the corner node and diffuse it across the network.

```{r}
#| label: lattice-raw

input_vec <- stats::setNames(5.0, "17")

obj <- diffuse_scores(obj, input = input_vec, method = "raw")
scores <- get_scores(obj)

head(scores[order(-input_1)])
```

### Visualising score decay

On a lattice, we expect scores to decay monotonically with shortest-path
distance from the source.

```{r}
#| label: lattice-decay-plot

dists <- distances(g, v = "17")
scores[, dist := as.integer(dists[1, node])]

decay_dt <- scores[, .(mean_score = mean(input_1)), by = dist]
setorder(decay_dt, dist)

ggplot(decay_dt, aes(x = dist, y = mean_score)) +
  geom_point(size = 2) +
  geom_line() +
  theme_bw() +
  labs(
    x = "Shortest-path distance from source",
    y = "Mean diffusion score",
    title = "Score decay on a 15x15 lattice (regularised Laplacian)"
  )
```

### Heatmap on the grid

Since this is a square lattice, we can visualise the scores as a heatmap.

```{r}
#| label: lattice-heatmap

scores[, row := (as.integer(node) - 1L) %/% 15L + 1L]
scores[, col := (as.integer(node) - 1L) %% 15L + 1L]

# square root the diffusion scores to make them nicer looking for plotting
# purposes
ggplot(scores, aes(x = col, y = row, fill = sqrt(input_1))) +
  geom_tile() +
  scale_fill_viridis_c(option = "mako") +
  coord_equal() +
  theme_minimal() +
  labs(
    fill = "Score",
    title = "Raw diffusion scores on a 15x15 lattice",
    subtitle = "Source node at bottom-left corner"
  )
```

We can see nicely how on this grid network the bottom left area of the graph
is getting the diffusion heat from the original source node.

### Comparing scoring methods

We can compare the three scoring approaches on the same input:

```{r}
#| label: lattice-methods

all_nodes <- V(g)$name
input_full <- stats::setNames(
  ifelse(all_nodes == "17", 1.0, 0.0),
  all_nodes
)

# z-score
obj_z <- build_kernel(
  DiffusionScores(g),
  kernel = "regularised_laplacian",
  strategy = "full",
  .verbose = FALSE
)
obj_z <- diffuse_scores(obj_z, input = input_full, method = "z")
z_scores <- get_scores(obj_z)

# monte carlo heatrank
obj_mc <- build_kernel(
  DiffusionScores(g),
  kernel = "regularised_laplacian",
  strategy = "full",
  .verbose = FALSE
)
obj_mc <- diffuse_scores(
  obj_mc,
  input = input_full,
  method = "mc",
  n_perm = 1000L,
  seed = 42L
)
mc_scores <- get_scores(obj_mc)

# combine
comparison <- merge(
  scores[, .(node, raw = input_1, dist)],
  z_scores[, .(node, z = input_1)],
  by = "node"
)
comparison <- merge(
  comparison,
  mc_scores[, .(node, heatrank = input_1)],
  by = "node"
)

head(comparison[order(-raw)])
```

```{r}
#| label: lattice-method-comparison

p1 <- ggplot(comparison, aes(x = dist, y = z)) +
  geom_jitter(width = 0.2, alpha = 0.5, size = 1) +
  theme_bw() +
  labs(x = "Distance from source", y = "Z-score")

p2 <- ggplot(comparison, aes(x = dist, y = heatrank)) +
  geom_jitter(width = 0.2, alpha = 0.5, size = 1) +
  theme_bw() +
  labs(x = "Distance from source", y = "Heatrank (lower = warmer)")

gridExtra::grid.arrange(p1, p2, ncol = 2)
```

## Biologically motivated example: scale-free network

Real biological networks (PPI, gene regulatory) tend to be scale-free: most
nodes have few connections, but a small number of hubs are highly connected.
We can simulate this with a Barabasi-Albert preferential attachment model.

### Generating the network

We will generate a random network first...

```{r}
#| label: scalefree-setup

set.seed(42)
g_sf <- sample_pa(n = 500, m = 3, directed = FALSE)
V(g_sf)$name <- paste0("gene_", seq_len(vcount(g_sf)))

# degree distribution
deg <- degree(g_sf)
hist(
  deg,
  breaks = 30,
  main = "Degree distribution (Barabasi-Albert, n=500)",
  xlab = "Degree",
  col = "steelblue",
  border = "white"
)
```

We can appreciate that most nodes have very few connections, with some having
the majority of edges in the network.

### Simulating a disease gene signal

We pick a handful of "seed genes" (mimicking known disease associations or gene
function) and diffuse their signal across the network to prioritise candidate 
genes.

```{r}
#| label: scalefree-seeds

# pick 5 seed genes: a mix of hubs and peripheral nodes
seed_genes <- c(
  names(sort(deg, decreasing = TRUE))[1:3], # top 3 hubs
  sample(names(deg[deg <= 3]), 2) # 2 peripheral nodes
)
cat("Seed genes:", paste(seed_genes, collapse = ", "), "\n")
cat("Seed degrees:", deg[seed_genes], "\n")
```

### Running diffusion

```{r}
#| label: scalefree-diffusion

obj_sf <- DiffusionScores(g_sf)

obj_sf <- build_kernel(
  obj_sf,
  kernel = "regularised_laplacian",
  kernel_params = params_kernel(sigma2 = 1.0, add_diag = 1.0),
  normalised = TRUE,
  strategy = "full",
  .verbose = TRUE
)

# input: seed genes get score 1, rest is background
all_genes <- V(g_sf)$name
input_sf <- stats::setNames(
  ifelse(all_genes %in% seed_genes, 1.0, 0.0),
  all_genes
)

obj_sf <- diffuse_scores(obj_sf, input = input_sf, method = "z")
results <- get_scores(obj_sf)
setorder(results, -input_1)

head(results, 20)
```

We can clearly appreciate that our seed genes have the most heat, followed by
the neighbours.

### Inspecting the top candidates

We can also investigate the top candidates that are not seed genes

```{r}
#| label: scalefree-results

results[, is_seed := node %in% seed_genes]
results[, rank := .I]
results[, degree := deg[node]]

# top 20 non-seed genes
top_candidates <- results[!(is_seed)][1:20]
top_candidates[, .(node, z_score = round(input_1, 3), degree)]
```

And explore the relationship of node degree with resulting Z-scores:

```{r}
#| label: scalefree-zscore-vs-degree

ggplot(results, aes(x = degree, y = input_1, fill = input_1, shape = is_seed)) +
  geom_point(alpha = 0.5, size = 2) +
  scale_fill_viridis_c(option = "mako") +
  scale_shape_manual(values = setNames(c(25, 21), c(TRUE, FALSE))) +
  theme_bw() +
  labs(
    x = "Node degree",
    y = "Z-score",
    fill = "Z-score",
    shape = "Seed gene:",
    title = "Diffusion z-scores vs degree (scale-free network)"
  )
```

### Comparing kernels

Different kernels emphasise different aspects of the network topology. The
diffusion kernel with a large bandwidth spreads signal more broadly, whilst
the regularised Laplacian is more conservative. For this synthetic example
however, we only tiny difference s

```{r}
#| label: scalefree-kernel-comparison

kernels <- c("regularised_laplacian", "diffusion", "pstep")
kernel_results <- list()

for (kt in kernels) {
  kp <- switch(
    kt,
    "diffusion" = params_kernel(sigma2 = 2.0),
    "pstep" = params_kernel(a = 3.0, p = 5L),
    params_kernel()
  )

  tmp <- build_kernel(
    DiffusionScores(g_sf),
    kernel = kt,
    kernel_params = kp,
    normalised = TRUE,
    strategy = "full",
    .verbose = FALSE
  )
  tmp <- diffuse_scores(tmp, input = input_sf, method = "raw")
  sc <- get_scores(tmp)
  sc[, kernel := kt]
  kernel_results[[kt]] <- sc
}

combined <- rbindlist(kernel_results)
combined[, is_seed := node %in% seed_genes]

# rank within each kernel
combined[, rank := frank(-input_1), by = kernel]

# overlap in top 20 across kernels
top_sets <- combined[
  rank <= 20 & is_seed == FALSE,
  .(genes = list(node)),
  by = kernel
]
cat("Top 20 non-seed genes per kernel:\n")
for (i in seq_len(nrow(top_sets))) {
  cat(sprintf(
    "  %s: %s\n",
    top_sets$kernel[i],
    paste(top_sets$genes[[i]], collapse = ", ")
  ))
}
```

```{r}
#| label: scalefree-kernel-rank-plot

# compare ranks across kernels
rank_wide <- dcast(
  combined[, .(node, kernel, rank)],
  node ~ kernel,
  value.var = "rank"
)

ggplot(rank_wide, aes(x = regularised_laplacian, y = diffusion)) +
  geom_point(alpha = 0.3, size = 1) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "red") +
  theme_bw() +
  labs(
    x = "Rank (regularised Laplacian)",
    y = "Rank (diffusion)",
    title = "Node ranking agreement between kernels"
  )
```

## Summary

The `DiffusionScores` workflow separates kernel construction from scoring,
so you can build the kernel once and run multiple diffusions with different
inputs or methods without recomputing the spectral decomposition. For large
networks (> 2000 nodes), the truncated strategy avoids the dense eigenvalue
decomposition entirely, using an iterative Lanczos solver that scales to
networks with tens of thousands of nodes.