---
title: "Embedding Drift via node2vec"
vignette: >
  %\VignetteIndexEntry{Embedding Drift via node2vec}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

# Context-aware differential embedding with node2vec

## Setup

```{r}
#| label: setup
library(genewalkR)
library(data.table)
library(magrittr)
library(ggplot2)
library(igraph)
```

## Background

This approach was proposed [Jassim, et al.](https://www.cell.com/cancer-cell/fulltext/S1535-6108(25)00255-7),
leveraging again node2vec under the hood. The core idea is that we can use
the embeddings generated by node2vec to understand across two networks that
share nodes (think co-expression network in pre-treatment vs post-treatment)
how the "context" (as surrogated by the embedding) has changed.

When we train node2vec on two different graphs independently, the resulting
embedding spaces are not directly comparable — the model initialises weights
randomly, so the same node can land anywhere in the vector space across runs.
To make the embeddings comparable, we apply an orthogonal Procrustes alignment
that rotates one embedding matrix onto the other using only nodes shared between
both graphs.

Once aligned, the cosine similarity between a node's vector in graph 1 and its
vector in graph 2 measures how much its local context has changed between the
two graphs. A similarity close to 1 means the node sits in a similar
neighbourhood in both graphs. A lower similarity signals that the topology
around that node has shifted.

Nodes that appear in only one graph are assigned sentinel values: `-1.1` for
nodes exclusive to graph 1 and `+1.1` for nodes exclusive to graph 2. These
fall outside the valid cosine similarity range of `[-1, 1]` and indicate
absence rather than a computed similarity.

## Synthetic data

To illustrate the method we use a synthetic graph pair with a known ground
truth. The two graphs share a common backbone of three communities but differ
in defined regions:

- **Community 1** is a stable negative control, topologically identical across
  both graphs.
- **Community 2** contains a hub node that is fully connected within the
  community in graph 1 but demoted to a single peripheral edge in graph 2.
- Two **bridge nodes** span communities 2 and 3 in graph 1 but are fully
  embedded within community 3 in graph 2.
- A small set of **exclusive nodes** appears in only one graph.
- A few cross-connections across the communities

We know in advance which nodes should show context drift and which should not,
making this a useful benchmark for the method.
```{r}
#| label: generate synthetic data
test_data <- differential_graph_test_data(
  n_stable = 50L,
  n_comm2 = 50L,
  n_comm3 = 50L,
  n_exclusive = 5L
)

# data_info holds the ground truth
head(test_data$data_info)
```

### Visualising the two graphs

Let's look at both graphs and highlight which nodes are expected to drift.
```{r}
#| label: igraph visualisation
#| fig-width: 12
#| fig-height: 6

# helper function for graph generation and plotting
make_graph <- function(edges, nodes, info) {
  g <- igraph::graph_from_data_frame(d = edges, directed = FALSE)

  node_names <- igraph::V(g)$name
  status <- info[match(node_names, info$node), node_status]
  differential <- info[match(node_names, info$node), is_differential]

  layout <- igraph::layout_with_kk(g)

  igraph::V(g)$x <- layout[, 1]
  igraph::V(g)$y <- layout[, 2]

  igraph::V(g)$color <- data.table::fcase(
    grepl("only", status)             , "grey80"  ,
    differential & status == "shared" , "#E74C3C" ,
    default = "#3498DB"
  )
  g
}

g1 <- make_graph(
  edges = test_data$g1_edges,
  nodes = test_data$g1_nodes,
  info = test_data$data_info
)

g2 <- make_graph(
  edges = test_data$g2_edges,
  nodes = test_data$g2_nodes,
  info = test_data$data_info
)

par(mfrow = c(1, 2))
plot(g1, vertex.label = NA, main = "Graph 1", vertex.size = 10)
plot(g2, vertex.label = NA, main = "Graph 2", vertex.size = 10)
par(mfrow = c(1, 1))
```

Nodes in red are those expected to show context drift. You can see the hub node
loses most of its connections in graph 2, and the bridge nodes shift from
spanning communities 2 and 3 to being embedded entirely within community 3.

## Running EmbedDrift

### Initialising the object

`EmbedDrift` takes the two edge tables and holds all intermediate results.
```{r}
#| label: initialise object

obj <- EmbedDrift(
  graph_dt_1 = test_data$g1_edges,
  graph_dt_2 = test_data$g2_edges
)

print(obj)
```

### Generating embeddings

We train node2vec independently on both graphs using identical hyperparameters.
Graph 1 serves as the reference for subsequent alignment. Given the small size
of this synthetic graph, a low embedding dimension of `8` is appropriate — 
using a larger dimension would over-parameterise relative to the number of nodes
and compress the signal we are trying to detect.
```{r}
#| label: generate embeddings

obj <- generate_initial_embeddings(
  object = obj,
  embd_dim = 8L,
  node2vec_params = params_node2vec(window_size = 2L),
  .verbose = TRUE
)

print(obj)
```

If you want to get the embeddings out, getters are provided:

```{r}
#| label: embeddrift getters

embd_ls <- get_embeddings(obj) # returns both matrices as a list. 

embd_ls$embd_1[1:5, 1:5]
```

### Calculating drift

`calculate_drift()` aligns the two embedding matrices via Procrustes and
computes per-node cosine similarities.
```{r}
#| label: calculate drift
obj <- calculate_drift(object = obj)

stats <- get_stats(obj)
stats
```

The results are ordered by cosine similarity ascending, so the nodes with the
greatest context drift appear first. Let's add the ground truth labels and see
whether the known differential nodes rank lower than the stable ones.
```{r}
#| label: evaluate results

results <- merge(
  stats,
  test_data$data_info[, .(node, is_differential)],
  by = "node"
)

results[node_status == "shared"] |>
  ggplot(aes(
    x = is_differential,
    y = cosine_similarity,
    fill = is_differential
  )) +
  geom_boxplot(width = 0.4, show.legend = FALSE, outlier.shape = NA) +
  geom_jitter(width = 0.1, shape = 21, size = 3, alpha = 0.6) +
  scale_fill_manual(values = c("FALSE" = "#3498DB", "TRUE" = "#E74C3C")) +
  scale_x_discrete(labels = c("FALSE" = "Stable", "TRUE" = "Differential")) +
  labs(
    fill = "Pos Ctr"
  ) +
  theme_minimal() +
  xlab("Pos Ctr") +
  ylab("Cosine sim") +
  ggtitle("Context drift by ground truth label")
```

Nodes with known topological changes (hub and bridge nodes) consistently show
lower cosine similarity than the stable community 1 nodes, confirming the
method recovers the planted signal. The exclusive nodes carry sentinel values
and are excluded from this comparison:

```{r}
#| label: sentinel values
# exclusive nodes carry fixed sentinel values
results[node_status != "shared", .(node, cosine_similarity, node_status)]
```