---
title: "Running GeneWalk"
vignette: >
  %\VignetteIndexEntry{How to run GeneWalk}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

# Exploring Gene Walk

This vignette will first show how Gene Walk behaves on synthetic data, what
assumptions are baked in and then we will move onto using it with real data.
If you want to understand the method in more detail, please check out 
[Ietswaart et al.](https://genomebiology.biomedcentral.com/articles/10.1186/s13059-021-02264-8).

## Loading in libraries

```{r}
#| label: setup
library(genewalkR)
library(ggplot2)
library(data.table)
library(magrittr)
```

## Synthetic data

### Exploring the data and initialising the object

Let's start with some synthetic data to understand how this works...
The synthetic GeneWalk network is designed testing and benchmarking, with
signal genes forming coherent, degree-matched graph neighbourhoods and noise
genes spanning multiple ontology subtrees at random.

```{r}
#| label: generate synthetic data
gene_walk_syn_data <- synthetic_genewalk_data()

str(gene_walk_syn_data)
```

The data contains everything we need to initialise a new GeneWalk:

- The full_data with the term ontology, gene to gene edges and gene to terms, 
mimicking relevant inputs.
- The genes to pathways for testing later.
- The gene identifiers including in the run. In this case, we have a set of 
"signal" genes that serve as a positive control and "noise genes" that are
randomly distributed.
- The term/pathway identifiers.

Let's initialise the class.

```{r}
#| label: initialise the class

genewalk_obj <- GeneWalk(
  graph_dt = gene_walk_syn_data$full_data,
  gene_to_pathway_dt = gene_walk_syn_data$gene_to_pathways,
  gene_ids = gene_walk_syn_data$gene_ids,
  pathway_ids = gene_walk_syn_data$pathway_ids
)

genewalk_obj
```

### Running gene walk on the synthetic data

This function here generates the node2vec embedding based on the network

```{r}
#| label: generate the first embeddings

genewalk_obj <- generate_initial_emb(
  genewalk_obj,
  node2vec_params = params_node2vec(),
  .verbose = TRUE
)
```

We need to generate a background distribution for testing purposes. This
will generate three random permuted networks as in the paper. These will be
used for statistical testing.

```{r}
#| label: generate the background distribution

genewalk_obj <- generate_permuted_emb(genewalk_obj, .verbose = TRUE)
```

We now need to compare the similarity of the Cosine similarities between
the gene embeddings to the pathway embeddings and check how often they are
larger than the ones from the background embedding, giving us the p-values.

```{r}
#| label: calculate the statistics

genewalk_obj <- calculate_genewalk_stats(
  genewalk_obj,
  .verbose = TRUE
)
```

Now we can extract the statistics:

```{r}
#| label: get the statistics

statistics <- get_stats(genewalk_obj)

head(statistics)
```

The different metrics in the data 
Let's explore the signal in the synthetic data a bit... What are we observing
... ?

```{r}
#| label: plot the similarities - synthetic

# add the labels for signal and noise
statistics[, signal := grepl("signal", gene)][, 
  signal := factor(signal, levels = c("TRUE", "FALSE"))
]

# plot the gene <> term similarities
ggplot(
  data = statistics,
  mapping = aes(x = similarity)
) + 
  geom_histogram(bins = 45) +
  facet_wrap(~ signal) +
  xlab("Cosine similarity") +
  ylab("Count")
```

As we can appreciate, the Cosine similarities between the genes and pathways
in the signal data set are all much higher compared to the FALSE ones. Let's 
look at the p-values

```{r}
#| label: plot the p-values - synthetic

ggplot(
  data = statistics,
  mapping = aes(x = avg_pval)
) + 
  geom_histogram(bins = 45) +
  facet_wrap(~ signal) +
  xlab("Cosine similarity") +
  ylab("Count")
```

Similar pattern. The `TRUE` signal ones are nearly all significant. Less so
for the noisy signal. The signal in the noise data is there, but this is 
expected due to the data. Let's look at genes passing the FDR thresholds

```{r}
table(
  grepl("signal", statistics$gene),
  statistics$avg_global_fdr < 0.05
)
```

Basically, all of our signal genes show significance associations between
gene and pathways; less so, for the noise genes. 

## Real data

Let's explore real data now. The package provides a builder factory to generate
the objects. Within the package, there is a DuckDB that contains 

**Network resources**

- The STRING network extracted from OpenTargets.
- The SIGNOR network extracted from OpenTargets.
- The Reactome gene to gene network extracted from OpenTargets.
- The Intact network extracted from OpenTargets.
- A combined network from the sources above, based on the approach from
  [Barrio-Hernandez, et al.](https://www.nature.com/articles/s41588-023-01327-9).

**Pathway terms**

- The Gene Ontology data extracted from the OBO files from the OBO foundry
  and OpenTargets.
- The Reactome pathway ontology and their gene to pathway associations from
  OpenTargets.

### Using the builder factory

Let's use the Gene Ontology and combined network for this example.

```{r}
#| label: initialise the factory

gw_factory <- GeneWalkGenerator$new()

gw_factory$add_pathways() # will add GO to the builder
gw_factory$add_ppi(source = "combined") # will add the combined one
gw_factory$build() # will load the data into the factory
```

The idea of the factory is to easily iterate through various bags of genes of
interest. Now let's use the factory to look specifically at the MYC target genes 
(provided in the package) and generate a GeneWalkNetwork for them.

```{r}
#| label: use the factory

data(myc_genes)

myc_gwn <- gw_factory$create_for_genes(genes = myc_genes$ensembl_gene)

myc_gwn
```

### Running gene walk on actual data

We can now use the same steps as above

```{r}
#| label: running gene walk on MYC

data(myc_genes)

myc_gwn <- gw_factory$create_for_genes(genes = myc_genes$ensembl_gene)

# this will take a while
myc_gwn <- generate_initial_emb(
  myc_gwn,
  node2vec_params = params_node2vec(n_epochs = 5L),
  .verbose = TRUE
)
# if you are on a many core machines, consider upping the num_workers here
# the default is 4L, but you can up this on your machine

# this even longer... go get coffee
myc_gwn <- generate_permuted_emb(
  myc_gwn,
  .verbose = TRUE
)

# let's calculate the statistics
myc_gwn <- calculate_genewalk_stats(
  myc_gwn,
  .verbose = TRUE
)

# let's extract the results
myc_gwn_res <- get_stats(myc_gwn)
```

### Exploring the results

Let's look at some of the ways to explore the data... Atm, we only have the
GO identifiers, let's add some names to it

```{r}
#| label: using the internal db

go_info <- get_gene_ontology_info()

go_translation <- setNames(
  go_info$go_name, go_info$go_id
)
```