---
title: "Metapath2vec"
date: last-modified
date-format: YYYY-MM-DD
vignette: >
  %\VignetteIndexEntry{Metapath2vec}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
author:
  Gregor Lueg
---

# Running metapath2vec

## Setup

```{r}
#| label: setup
library(genewalkR)
library(data.table)
library(magrittr)
library(ggplot2)
```

## Background

Biological networks are rarely homogeneous. Genes interact with genes, genes
sit in pathways, pathways sit in a hierarchy. Plain node2vec throws all of that
into one pot and walks wherever the edges go. metapath2vec
([Dong, Chawla and Swami](https://ericdongyx.github.io/papers/KDD17-dong-chawla-swami-metapath2vec.pdf))
keeps the node types and forces every walk to follow a schema over them, a
so-called metapath. `gene-pathway-gene` means: from a gene, hop to one of its
pathways, then to another gene in that pathway, and repeat. The skip-gram
model on top is the node2vec one; only the walks change.

Two things to know about the implementation:

- The walks are first-order. The next hop depends only on the current node
  and the position in the schema, so there's no `p` and `q`, and no
  `(previous, current)` transition table that blows up memory on dense graphs.
- `metapath_plus = TRUE` gives you metapath2vec++. Negative samples are then
  drawn from the context node's own type, so a gene is contrasted against
  other genes and a pathway against other pathways.

## Building a typed Reactome graph

The package ships a DuckDB with Reactome, and it has everything we need: gene
to gene interactions, gene to pathway membership and the pathway hierarchy.
We use two node types, `gene` and `pathway`.

```{r}
#| label: load reactome
gene_gene <- get_interactions_reactome()[, .(
  from = as.character(from),
  to = as.character(to)
)]
gene_pathway <- get_gene_to_reactome()[, .(
  from = as.character(from),
  to = as.character(to)
)]
pathway_pathway <- get_reactome_hierarchy("child_of")[, .(
  from = as.character(from),
  to = as.character(to)
)]

edges <- rbind(gene_gene, gene_pathway, pathway_pathway) %>%
  unique() %>%
  .[from != to]

nodes <- rbind(
  data.table(
    id = unique(c(gene_gene$from, gene_gene$to, gene_pathway$from)),
    type = "gene"
  ),
  data.table(
    id = unique(c(gene_pathway$to, pathway_pathway$from, pathway_pathway$to)),
    type = "pathway"
  )
) %>%
  unique(by = "id")

nodes[, .N, by = type]
nrow(edges)
```

metapath2vec wants two tables: the edges with `from` and `to` (and optionally
`weight`), and the nodes with `id` and `type`. Every edge endpoint needs to
be in the node table. For readable output later on, we also pull the gene
symbols and pathway names.

```{r}
#| label: labels
node_labels <- rbind(
  get_gene_info()[, .(
    id = as.character(ensembl_id),
    label = as.character(symbol)
  )],
  get_reactome_info()[, .(
    id = as.character(reactome_id),
    label = as.character(reactome_name)
  )]
)
```

## Check your walk statistics first

A schema that doesn't fit the graph still gives you vectors. They're just
garbage. Every run records what happened during walk generation in the
`walk_stats` attribute, so look at it! Here's a schema that sounds sensible,
from a gene to its pathway, up or down the hierarchy, and back to a gene:

```{r}
#| label: bad schema
bad_schema <- metapath2vec(
  graph_dt = edges,
  node_dt = nodes,
  metapath = c("gene", "pathway", "pathway", "gene"),
  metapath2vec_params = params_metapath2vec(
    walks_per_node = 5L,
    n_epochs = 1L
  ),
  .verbose = FALSE
)

bad_stats <- attr(bad_schema, "walk_stats")
str(bad_stats)
```

`r round(100 * bad_stats$truncated / bad_stats$attempted, 1)`% of the walks
dead-end before they reach the requested length, and the mean realised length
is `r round(bad_stats$mean_length, 1)` instead of `r bad_stats$walk_length`.
Genes are mostly annotated to the leaves of the Reactome hierarchy, so the
parent pathway you land on has no genes of its own to hop to. The schema is
fine on paper; it just doesn't match how this graph is wired.

## Gene embeddings

`gene-pathway-gene` is the canonical choice. It ignores the interaction edges
completely, so the embedding encodes shared pathway membership and nothing
else. We run it as plain metapath2vec and as metapath2vec++, and add node2vec
over the same edge table as the baseline.

```{r}
#| label: gene embeddings
walk_params <- params_metapath2vec(
  walks_per_node = 20L,
  walk_length = 40L,
  n_epochs = 10L
)

gpg <- metapath2vec(
  graph_dt = edges,
  node_dt = nodes,
  metapath = c("gene", "pathway", "gene"),
  embd_dim = 32L,
  metapath2vec_params = walk_params,
  .verbose = FALSE
)

str(attr(gpg, "walk_stats"))

gpg_plus <- metapath2vec(
  graph_dt = edges,
  node_dt = nodes,
  metapath = c("gene", "pathway", "gene"),
  embd_dim = 32L,
  metapath2vec_params = walk_params,
  metapath_plus = TRUE,
  .verbose = FALSE
)

n2v <- node2vec(
  graph_dt = edges,
  embd_dim = 32L,
  node2vec_params = params_node2vec(
    walks_per_node = 20L,
    walk_length = 40L,
    n_epochs = 10L
  ),
  .verbose = FALSE
)
```

This time the walk statistics look healthy: hardly any walk is truncated.

### Do genes cluster by biology?

To score the embeddings, we need a label per gene. Reactome has
`r length(setdiff(nodes[type == "pathway", id], pathway_pathway$from))`
top-level pathways (signal transduction, metabolism, immune system, ...). We
walk the `child_of` edges up to them, and keep only the genes whose pathways
all lead to a single top-level pathway, so the label is unambiguous.

```{r}
#| label: top-level labels
parents <- pathway_pathway[, .(child = from, parent = to)]
roots <- setdiff(nodes[type == "pathway", id], parents$child)

# climb the hierarchy until every pathway has reached a root
ancestry <- data.table(pathway = nodes[type == "pathway", id])
ancestry[, top_level := pathway]
while (!all(ancestry$top_level %in% roots)) {
  ancestry <- merge(
    ancestry,
    parents,
    by.x = "top_level",
    by.y = "child",
    all.x = TRUE,
    allow.cartesian = TRUE
  ) %>%
    .[!is.na(parent), top_level := parent] %>%
    .[, parent := NULL] %>%
    unique()
}

gene_top_level <- merge(
  gene_pathway[, .(gene = from, pathway = to)],
  ancestry,
  by = "pathway"
)[, .(gene, top_level)] %>%
  unique()

gene_top_level <- gene_top_level[, if (.N == 1L) .SD, by = gene]

nrow(gene_top_level)
```

The score is kNN purity: for each gene, the share of its 10 nearest
neighbours (cosine) with the same top-level pathway. To keep this quick we
sample 2,000 of the labelled genes.

```{r}
#| label: knn purity
knn_purity <- function(embd, labels, k = 10L) {
  x <- embd[labels$gene, ]
  x <- x / sqrt(rowSums(x^2))
  sim <- tcrossprod(x)
  diag(sim) <- -Inf
  nn <- t(apply(sim, 1, \(s) order(s, decreasing = TRUE)[seq_len(k)]))
  mean(matrix(labels$top_level[nn], ncol = k) == labels$top_level)
}

set.seed(42L)
eval_genes <- gene_top_level[sample(.N, 2000L)]

purity <- data.table(
  method = c(
    "node2vec",
    "metapath2vec (gene-pathway-gene)",
    "metapath2vec++ (gene-pathway-gene)",
    "random"
  ),
  knn_purity = c(
    knn_purity(n2v, eval_genes),
    knn_purity(gpg, eval_genes),
    knn_purity(gpg_plus, eval_genes),
    sum((eval_genes[, .N, by = top_level]$N / nrow(eval_genes))^2)
  )
)

purity
```

All three are miles above random and within a few points of each other. The
order shuffles between runs, since Hogwild training isn't deterministic. That
node2vec keeps up is no surprise: it sees the pathway edges as well, plus the
interactions and the hierarchy, and the label is derived from the hierarchy.
metapath2vec gets there from membership alone. metapath2vec++ barely moves the
needle on this graph, since there are only two types and the gene side
dominates the walks anyway. So what's the point? You get to pick what
"similar" means. With node2vec, you get whatever the graph mix gives you.

### Nearest neighbours

Genes and pathways live in the same space, so we can ask one embedding for
both. Here are the neighbours of TP53:

```{r}
#| label: tp53 neighbours
nearest <- function(embd, query, n = 10L) {
  x <- embd / sqrt(rowSums(embd^2))
  sim <- drop(x %*% x[query, ])
  top <- order(sim, decreasing = TRUE)[seq_len(n) + 1L]
  data.table(id = rownames(embd)[top], cosine = round(sim[top], 3)) %>%
    merge(nodes, by = "id", sort = FALSE) %>%
    merge(node_labels, by = "id", all.x = TRUE, sort = FALSE)
}

tp53 <- node_labels[label == "TP53", id][1]

nearest(gpg, tp53, n = 15L)
```

## Pathway embeddings

Flip the schema to `pathway-gene-pathway` and the walks start on pathways.
Two pathways end up close when they share genes, which is a smooth version of
the Jaccard overlap you'd otherwise compute by hand. And it's fast, because
there are far fewer pathways than genes to start walks from.

```{r}
#| label: pathway embeddings
pgp <- metapath2vec(
  graph_dt = edges,
  node_dt = nodes,
  metapath = c("pathway", "gene", "pathway"),
  embd_dim = 32L,
  metapath2vec_params = walk_params,
  .verbose = FALSE
)

str(attr(pgp, "walk_stats"))
```

About a fifth of the walks get dropped straight away. They start on the
higher-level pathways with no directly annotated genes, same story as above.
No walk ever touches those pathways, so all they'd have is their random
initialisation. The same goes for genes with interactions only, since this
schema never uses the gene-gene edges. `metapath2vec()` removes such nodes by
default and lists them in the `unvisited_nodes` attribute (set
`filter_unvisited = FALSE` if you want the rows anyway):

```{r}
#| label: unvisited nodes
unvisited <- attr(pgp, "unvisited_nodes")

nodes[id %in% unvisited, .N, by = type]
dim(pgp)

pathway_ids <- intersect(rownames(pgp), nodes[type == "pathway", id])
```

A pathway's closest neighbours are its own member genes, so for pathway to
pathway similarity we search the pathway rows only. Neighbours of
TNFR1-induced proapoptotic signalling:

```{r}
#| label: tnfr1 neighbours
tnfr1 <- node_labels[label == "TNFR1-induced proapoptotic signaling", id][1]

nearest(pgp[pathway_ids, ], tnfr1)
```

A PCA over the pathway rows, coloured by the eight largest top-level
pathways:

```{r}
#| label: pathway pca
pca_results <- prcomp(pgp[pathway_ids, ])

top_levels <- ancestry[, .N, by = top_level][order(-N)][1:8, top_level]

pca_dt <- as.data.table(pca_results$x[, 1:2], keep.rownames = "pathway") %>%
  merge(ancestry, by = "pathway") %>%
  .[, if (.N == 1L) .SD, by = pathway] %>%
  .[top_level %in% top_levels] %>%
  merge(
    node_labels[, .(top_level = id, top_level_name = label)],
    by = "top_level"
  )

ggplot(data = pca_dt, mapping = aes(x = PC1, y = PC2)) +
  geom_point(
    mapping = aes(color = top_level_name),
    size = 1,
    alpha = 0.7
  ) +
  theme_bw() +
  labs(color = "Top-level pathway:")
```

Two components out of 32 are a blunt view, but the broad layout is there:
metabolism on one side, signal transduction on the other, immune system and
developmental biology pulling apart along PC2. Disease is all over the place,
which is fair: Reactome's disease branch reuses the genes of every other
branch.

## Where to go from here

The walk engine takes up to 256 node types, so adding Gene Ontology as a third
type is a table join away: `gene-go-gene` next to `gene-pathway-gene` gives you
two views of function you can compare. On the roadmap: several metapaths in
one run, which the original paper also discusses and which avoids having to
pick a single schema up front.
