---
title: "Node2vec"
vignette: >
  %\VignetteIndexEntry{Node2vec}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
knitr:
  opts_chunk:
    collapse: true
    comment: '#>'
---

# Running node2vec

## Loading in libraries

```{r}
#| label: setup
library(genewalkR)
library(data.table)
library(magrittr)
library(ggplot2)
```

## Synthetic data

genewalkR provides quite a few synthetic graph data sets for testing and
exploration purposes, namely:

- **Barbell graph**: a graph of two distinct clusters
- **Cavemen graph**: a graph with distinct communities with a low probability of
  connection between them.
- **Stochastic blocks**: a more densely connected graph with communities; 
  however, more cross-edges compared to the previous graphs.

### Barbell graph

Let's load in the barbell graph.

```{r}
#| label: loading in barbell data

# this creates a random data set with 20 nodes per barbell cluster
barbell_data <- node2vec_test_data(
  test_type = "barbell",
  n_nodes_per_cluster = 20L
)
```

#### Exploring the data

Let's transform this into an igraph first and just look at the data like this

```{r}
#| label: generate igraph from barbell data

g <- igraph::graph_from_data_frame(d = barbell_data$edges, directed = FALSE)

plot(g, vertex.label = NA)
```

#### Running node2vec

You can appreciate a nice barbell graph with two distinct groups. Now let's
run node2vec on this data. The node2vec implementation is based on
[node2vec-rs](https://github.com/GregorLueg/node2vec-rs), a Rust-based
implementation of node2vec. In its default setting, we use a HIGHLY optimised
version that runs on the CPU with SIMD-acceleration and a tolerated race
condition (undefined behaviour in low level terms) that makes this very fast.
The issue with this is that the underlying embeddings will not be deterministic;
however, the structure of the underlying graph will be identified.

```{r}
#| label: run node2vec

# run node2vec
barbell_node2vec_results <- node2vec(
  graph_dt = barbell_data$edges,
  .verbose = TRUE
)

# the parameters are controlled via a supplied parameter list, see code here
str(params_node2vec())

# this list can be supplied and tells node2vec how to run
```

If you want more details on how node2vec actually works, please, check the
original paper from [Grover and Leskovec](https://arxiv.org/abs/1607.00653).
Let's project that data onto two dimensions via PCA and visualise the results

```{r}
#| label: node2vec pca

# run pca on the embeddings
pca_results <- prcomp(barbell_node2vec_results, scale. = TRUE)

pca_dt <- as.data.table(pca_results$x[, 1:2], keep.rownames = "node") %>%
  merge(., barbell_data$node_labels, by = "node")

pca_dt[, cluster := factor(cluster)]

ggplot(data = pca_dt, mapping = aes(x = PC1, y = PC2)) +
  geom_point(mapping = aes(color = cluster)) 
```

As you can appreciate, node2vec embeddings nicely identify the two groups
in the underlying graph. 

## Other graphs

Let's look at some other synthetic graphs...

### Cavemen graph

Visualisation of the graph structure in igraph first:

```{r}
#| label: cavemen igraph

# generate the cavemen
caveman_data <- node2vec_test_data(
  test_type = "cavemen", n_clusters = 5L, p_between = 0.5
)

g <- igraph::graph_from_data_frame(d = caveman_data$edges, directed = FALSE)

plot(g, vertex.label = NA)
```

Let's check out node2vec on this graph

```{r}
#| label: cavemen node2vec visualisation

# run node2vec
cavemen_node2vec_results <- node2vec(
  graph_dt = caveman_data$edges,
  .verbose = TRUE
)

# run pca on the embeddings
pca_results <- prcomp(cavemen_node2vec_results, scale. = TRUE)

pca_dt <- as.data.table(pca_results$x[, 1:2], keep.rownames = "node") %>%
  merge(., caveman_data$node_labels, by = "node")

pca_dt[, cluster := factor(cluster)]

ggplot(data = pca_dt, mapping = aes(x = PC1, y = PC2)) +
  geom_point(mapping = aes(color = cluster)) 
```

### Stochastic blocks 

```{r}
#| label: stochastic block igraph

# generate the stochastic block graph
stochastic_block_data <- node2vec_test_data(
  test_type = "stochastic_block", n_clusters = 3L, p_between = 0.05
)

g <- igraph::graph_from_data_frame(
  d = stochastic_block_data$edges, 
  directed = FALSE
)

plot(g, vertex.label = NA)
```

Let's check out node2vec on this graph

```{r}
#| label: stochastic block node2vec visualisation

# run node2vec
stochastic_blocks_node2vec_results <- node2vec(
  graph_dt = stochastic_block_data$edges,
  .verbose = TRUE
)

# run pca on the embeddings
pca_results <- prcomp(stochastic_blocks_node2vec_results, scale. = TRUE)

pca_dt <- as.data.table(pca_results$x[, 1:2], keep.rownames = "node") %>%
  merge(., stochastic_block_data$node_labels, by = "node")

pca_dt[, cluster := factor(cluster)]

ggplot(data = pca_dt, mapping = aes(x = PC1, y = PC2)) +
  geom_point(mapping = aes(color = cluster)) 
```

In conclusion, node2vec generates an embedding of the underlying network 
structure that you can then use to visualise the data or do additional analysis
(supervised tasks or like in this package statistical testing for gene <>
pathway associations in a bag of genes).